---
title: "Grant Opportunities Related to `r params$topic`" # Update the title if desired
subtitle: "" # Add a subtitle if desired, e.g., "Madison Research Club February 2026"
date: today # You can change the date to a specific date if you want, e.g., "February 11, 2026"
date-format: long
language:
  title-block-published: "Date"
author: 
  - name: "[Office of Research Development](https://www.jmu.edu/research/development/index.shtml)"
    email: researchdevelopment@jmu.edu
    affiliations: 
      - name: "REDI Division\n\nJames Madison University"
brand: _brand.yml
format: 
  typst:
    template: ../html/jmu_grant_prospecting.typ
editor: source
warning: false
echo: false
embed-resources: true
params:
  topic: "A topic" # Update with the desired topic for the grant opportunities
  program_ids: [500, 501, 100] # Update with relevant program IDs for the topic (program_id column)
  date_cutoff: "2026-04-01" # "YYYY-MM-DD" Update with the desired cutoff date for upcoming deadlines
  url: "https://docs.google.com/spreadsheets/..." # Update with the URL of the Google Sheet containing the grants data
---


```{r}
#| label: setup
#| include: false

# Auto-install any missing packages
required_packages <- c("tidyverse", "lubridate", "tinytable", "googlesheets4")

missing <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if (length(missing) > 0) install.packages(missing)

library(tidyverse)
library(lubridate)
library(tinytable)
```

# Summary


The following are some upcoming grant opportunities related to `r params$topic`.

Feel free to reach out to the [Office of Research Development](https://www.jmu.edu/research/development/index.shtml) for assistance with identifying more specific funding opportunities, developing proposals, and finding collaborators.

# Current Opportunities

```{r}

# grants <- read_csv("data/coe_2025_grants.csv")

googlesheets4::gs4_auth(email = TRUE) 

grants <- googlesheets4::read_sheet(params$url) |>
  select(program, funder, url, deadline, amount, synopsis, notes, program_id, discipline) |>
  filter(program_id %in% c(params$program_ids)) 

# grants <- read_csv("data/grants.csv") |>
#   select(program, funder, url, deadline, amount, synopsis, notes, program_id) |>
#   filter(program_id %in% c(params$program_ids)) 


```


```{r}
# ── Typst-safe helpers ────────────────────────────────────────────────────────

# Escape characters that Typst treats as special syntax.
# $ opens math mode; \ is an escape character itself.
typst_escape <- function(x) {
  x <- gsub("\\\\", "\\\\\\\\", x)   # backslash first (must be before $)
  x <- gsub("\\$",  "\\\\$",   x)    # dollar sign  → \$
  x
}

# Format a grant amount and escape it for Typst.
format_amount <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x)) return("")
  x_char <- as.character(x)
  if (x_char %in% c("NULL", "Not Specified", "")) return(x_char)
  # Range / level strings — escape any $ already present then return
  if (grepl("Level|to", x_char)) return(typst_escape(x_char))
  # Already formatted with a leading $ — escape and return
  if (grepl("^\\$", x_char)) return(typst_escape(x_char))
  # Plain number — format with commas and a Typst-safe dollar sign
  x_num <- suppressWarnings(as.numeric(x_char))
  if (!is.na(x_num)) {
    return(paste0("\\$", format(x_num, big.mark = ",", scientific = FALSE)))
  }
  typst_escape(x_char)
}

# Build a red Typst hyperlink (uses the #link() built-in).
typst_link <- function(url, label) {
  paste0('#link("', url, '")[', label, ']')
}

# Build a purple Typst hyperlink (uses the #plink() helper from jmu.typ).
typst_purple_link <- function(url, label) {
  paste0('#plink("', url, '")[', label, ']')
}

# Prepare a data frame for tt(): escape text columns, format amounts,
# build hyperlinked program column, then rename for display.
prepare_grants <- function(df, purple_links = FALSE) {
  link_fn <- if (purple_links) typst_purple_link else typst_link

  df |>
    mutate(
      # Hyperlinked program name
      Program  = link_fn(url, program),
      # Escape free-text columns so stray $ / \ don't break Typst
      Sponsor  = typst_escape(as.character(funder)),
      Deadline = as.character(deadline),
      Amount   = map_chr(amount, format_amount),
      Synopsis = typst_escape(as.character(synopsis))
    ) |>
    select(Program, Sponsor, Deadline, Amount, Synopsis)
}

# Render a grants data frame as a Typst-ready tinytable.
grants_table <- function(df, purple_links = FALSE) {
  prepared <- prepare_grants(df, purple_links = purple_links)
  tbl <- prepared |>
    tt(
      escape = FALSE,
      width  = c(0.18, 0.15, 0.12, 0.12, 0.43)
    )

  if (nrow(prepared) > 1) {
    tbl <- tbl |>
      style_tt(
        i          = seq_len(nrow(prepared) - 1),
        line       = "b",
        line_color = "#cccccc",
        line_width = 0.05
      )
  }

  tbl
}
```


```{r}
#| label: current-filter

grants_current <- grants |>
  mutate(deadline = map_chr(deadline, ~ifelse(is.null(.x), "Rolling", .x))) |>
  mutate(deadline_parsed = mdy(deadline)) |>
  filter(deadline_parsed > as.Date(params$date_cutoff) | is.na(deadline_parsed)) |>
  filter(discipline != "Research Development") |>
  arrange(deadline_parsed) |>
  select(-deadline_parsed, -discipline)
```

```{r}
#| label: current-table

grants_table(grants_current)
```

# Grants to Monitor

```{r}
#| label: past-filter

grants_past <- grants |>
  mutate(deadline = map_chr(deadline, ~ifelse(is.null(.x), "Rolling", .x))) |>
  mutate(deadline_parsed = mdy(deadline)) |>
  filter(deadline_parsed <= as.Date(params$date_cutoff) | discipline == "Research Development") |>
  arrange(deadline_parsed) |>
  select(-deadline_parsed, -discipline)
```

```{r}
#| label: past-table

# Pass purple_links = TRUE to render the Program column in JMU purple
# instead of the default red. Requires #plink() defined in jmu.typ.

grants_past <- grants |>
  mutate(deadline = map_chr(deadline, ~ifelse(is.null(.x), "Rolling", .x))) |>
  mutate(deadline_parsed = mdy(deadline)) |>
  filter(deadline_parsed <= as.Date(params$date_cutoff)) |>
  arrange(deadline_parsed) |>
  select(-deadline_parsed)
```

# Contact

Office of Research Development

**Email:** [researchdevelopment@jmu.edu](mailto:researchdevelopment@jmu.edu)

